Model-based Clustering for Neural Populations

### 2021-09-15 documents/cluster_neural_population.tex
documentation-Poisson MCMC

@@

DETOUR: we can write things in the form of MV-(Poisson)-GLM as follows (in cluster j with \(n_j\) observations/ neurons, with corresponding loading be \(\mathbf{d}_{j}\) and \(\mathbf{C}_{j}\)):

\[\log\begin{pmatrix}
	\lambda_{11} & \cdots & \lambda_{1n_{j}} \\
	\vdots & \ddots & \vdots \\
	\lambda_{T1} & \cdots & \lambda_{Tn_{j}} \\
\end{pmatrix} = \begin{pmatrix}
	1 & \mathbf{x}_{1}^{'} \\
	\vdots & \vdots \\
	1 & \mathbf{x}_{T}^{'} \\
\end{pmatrix}\begin{pmatrix}
	\mathbf{d}_{j}^{'} \\
	\mathbf{C}_{j}^{'} \\
\end{pmatrix} + \mathbf{E}_j\]

We can further organize things column-wise. let \(\begin{pmatrix}
	1 & \mathbf{x}_{1}^{'} \\
	\vdots & \vdots \\
	1 & \mathbf{x}_{T}^{'} \\
\end{pmatrix} = \mathbf{G}\), then the design matrix and corresponding log-mean response are \(\begin{pmatrix}
\mathbf{G} & \mathbf{0} & \cdots & \mathbf{0} \\
\mathbf{0} & \mathbf{G} & \cdots & \mathbf{0} \\
\vdots & \vdots & \ddots & \vdots \\
\mathbf{0} & \cdots & \cdots & \mathbf{G} \\
\end{pmatrix}\) and \({\log\begin{pmatrix}
	\lambda_{11}^{(j)} & \cdots & \lambda_{T1}^{(j)} & \lambda_{12}^{(j)} & \cdots & \lambda_{T2}^{(j)} & \cdots & \lambda_{Tn_{j}}^{(j)} \\
\end{pmatrix}}'\).

However, when I use the Laplace approximation and NR, because of the block-diagonal structure of design matrix, the error is always independent observation by observation. This is equivalent to observation-wise update as above.

\textbf{Maybe I did something wrong?}
The point to say this is to make loading within each cluster depends on each other. Check these two later(\href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}).

Now I do it in another way: make the prior cluster-dependent...

@@

\section{Simulations}

@@

\subsubsection{Labeled Data: No Clustering}
The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}. After running 2000 iterations, the trace of the sample is shown in Figure \ref{lds norms}.
\begin{figure}[h!]
	\centering
	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
	\label{lds norms}
\end{figure}\\

Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
	}
	\caption{LDS sample with labels}
	\label{fig:LDS labeled}
\end{figure}

Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).

@@

\textbf{Think more on it.} Maybe we can do MV-(Poisson)-GLM, to make all loading within cluster correlated.

@@

This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
\begin{figure}[h!]
	\centering
	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
	\label{noA norms}
\end{figure}\\

Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
	}
	\caption{LDS sample with labels}
	\label{fig:noA labeled}
\end{figure}

This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}

\subsubsection{Unlabeled Data: Clustering}
Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}. The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.

@@

\section{TODO}
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
	Think about MV-(Poisson)-GLM. Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
	\item
	After resolving (1), see if the MCMC is still trapped in local optima sometimes. Now, the chain seems will get stuck in local mode. Maybe updating the loading as a whole will remedy the problem a bit.
	\item
	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
	\item
	improve DPMM. In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)). There are several improvements, e.g. \href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}. Check them later.
	\item
	After all of them are resolved, switch to GP version

>>>>

\textbf{Step Further:}\\
Using normal approximation may not be accurate enough, and this may lead to a bad mixing. We can step further to use the Metropolis-Hasting, based on the normal approximated variance. To be specific, the proposal distribution is \(Q(\mathbf{x}^{*}|\mathbf{x}^{(s)}) \sim N(\mathbf{x}^{(s)}, \alpha\mathbf{\Sigma}_{\mathbf{x}})\), where \(\alpha\) is a scalar to make the acceptance rate to be 0.4 to 0.6. According to \href{https://projecteuclid.org/journals/statistical-science/volume-16/issue-4/Optimal-scaling-for-various-Metropolis-Hastings-algorithms/10.1214/ss/1015346320.full}{Roberts and Rosenthal, 2001}. For high dimensional MH, the optimal proposal is \(N(x, 2.38^2\Sigma/d)\). See details \& experiments in Section \ref{sim}: Simulations.

@@

%DETOUR: we can write things in the form of MV-(Poisson)-GLM as follows (in cluster j with \(n_j\) observations/ neurons, with corresponding loading be \(\mathbf{d}_{j}\) and \(\mathbf{C}_{j}\)):
%
%\[\log\begin{pmatrix}
%	\lambda_{11} & \cdots & \lambda_{1n_{j}} \\
%	\vdots & \ddots & \vdots \\
%	\lambda_{T1} & \cdots & \lambda_{Tn_{j}} \\
%\end{pmatrix} = \begin{pmatrix}
%	1 & \mathbf{x}_{1}^{'} \\
%	\vdots & \vdots \\
%	1 & \mathbf{x}_{T}^{'} \\
%\end{pmatrix}\begin{pmatrix}
%	\mathbf{d}_{j}^{'} \\
%	\mathbf{C}_{j}^{'} \\
%\end{pmatrix} + \mathbf{E}_j\]
%
%We can further organize things column-wise. let \(\begin{pmatrix}
%	1 & \mathbf{x}_{1}^{'} \\
%	\vdots & \vdots \\
%	1 & \mathbf{x}_{T}^{'} \\
%\end{pmatrix} = \mathbf{G}\), then the design matrix and corresponding log-mean response are \(\begin{pmatrix}
%\mathbf{G} & \mathbf{0} & \cdots & \mathbf{0} \\
%\mathbf{0} & \mathbf{G} & \cdots & \mathbf{0} \\
%\vdots & \vdots & \ddots & \vdots \\
%\mathbf{0} & \cdots & \cdots & \mathbf{G} \\
%\end{pmatrix}\) and \({\log\begin{pmatrix}
%	\lambda_{11}^{(j)} & \cdots & \lambda_{T1}^{(j)} & \lambda_{12}^{(j)} & \cdots & \lambda_{T2}^{(j)} & \cdots & \lambda_{Tn_{j}}^{(j)} \\
%\end{pmatrix}}'\).
%
%However, when I use the Laplace approximation and NR, because of the block-diagonal structure of design matrix, the error is always independent observation by observation. This is equivalent to observation-wise update as above.
%
%\textbf{Maybe I did something wrong?}
%The point to say this is to make loading within each cluster depends on each other. Check these two later(\href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}).
%
%Now I do it in another way: make the prior cluster-dependent...

\textbf{Step further}:
Again, use MH to sample the posterior, based on the normal approximated variance. Since the dimension of d and C is not large (just d = 3) in the simulation, I didn't use the optimal scalar yet. \textbf{Modify it later}.

@@

\section{Simulations}\label{sim}

@@

\subsubsection{Labeled Data: No Clustering}
The code for fitting is \href{https://github.com/weigcdsb/state-space-clustering/blob/main/LDS/blkDiag/lds_sample_DP_blkDiag.m}{here}.

Here, I first compare 10k iterations for three methods (consider high dimension of \(\mathbf{x}_t\)): (1) both \(\mathbf{x}_t\) and loading (\(\mathbf{d}\) and \(\mathbf{C}\)) are updated by normal approximation. (2) update \(\mathbf{x}_t\) by normal approximation, but update loading by MH. (3) update both \(\mathbf{x}_t\) and loading by MH.

The fitting results are average of iteration 5k to 10k.

\textbf{Update \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by normal approximation}

The trace plots are in Figure \ref{double N: trace}
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleN.png}}%
		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleN.png}}%
	}
	\caption{both Normal Approx., trace plots}
	\label{double N: trace}
\end{figure}

Well, the chain hasn't achieve the stationary distribution. So there's no need to see further results. But to show the improvement and for comparison, I still show the fitting of overall firing rate, latents and dynamics in Figure \ref{double N: fit}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleN.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleN.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleN.png}}%
	}
	\caption{both Normal Approx., fit}
	\label{double N: fit}
\end{figure}

\textbf{Update \(\mathbf{x}_t\) by normal approximation and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}

By using the normal approximated posterior variance as the proposal variance, the acceptance rates are around 0.45 (Since there are 30 neurons, there are 30 acceptance rates).
The trace plots are in Figure \ref{dc MH: trace}
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_dcMH.png}}%
		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_dcMH.png}}%
	}
	\caption{latent Normal Approx. + loading MH, trace plots}
	\label{dc MH: trace}
\end{figure}

Looks much better, although still not perfect yet. The fitting of overall firing rate, latents and dynamics are in Figure \ref{dc MH: fit}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_dcMH.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_dcMH.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_dcMH.png}}%
	}
	\caption{latent Normal Approx. + loading MH, fit}
	\label{dc MH: fit}
\end{figure}

See, nearly perfect (especially) for latent.

\textbf{Update both \(\mathbf{x}_t\) and (\(\mathbf{d}\), \(\mathbf{C}\)) by MH}

The acceptance rate (by using the optimal scalar) for latent is 0.6227, which is a bit high (but acceptable). Again, the acceptance rates for loading are around 0.45.

The trace plots are in Figure \ref{double MH: trace}
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_doubleMH.png}}%
		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_doubleMH.png}}%
	}
	\caption{both MH, trace plots}
	\label{double MH: trace}
\end{figure}

The trace for dynamics is terrible... Maybe this is caused by high dimension of latent (\(d=6000\) in this case).

The fitting of overall firing rate, latents and dynamics are in Figure \ref{double MH: fit}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_doubleMH.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_doubleMH.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_doubleMH.png}}%
	}
	\caption{both MH, fit}
	\label{double MH: fit}
\end{figure}

OK, the fitting sucks...

One natural idea is to combine MH and normal approximation in latent: at first few steps, use normal approximation at conditional posterior mode to bring things into somewhere close to mode quickly. Then use MH to stabilize everything.

In the following, I still sample loading by MH. But for latent, the first half iteration is sampled by normal approximation, while the remainder is sampled from MH.

The red line marks when I switch the sampling method for \(\mathbf{x}_t\).

The trace plots are in Figure \ref{half MH: trace}
\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[tracd of latent]{\label{fig:a}\includegraphics[width=0.55\textwidth]{xnorm_halfMH.png}}%
		\subfigure[trace of loading and dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{trace_halfMH.png}}%
	}
	\caption{half MH for latent, trace plots}
	\label{half MH: trace}
\end{figure}

The fitting of overall firing rate, latents and dynamics are in Figure \ref{half MH: fit}.

\begin{figure}[h!]
	\makebox[\linewidth][c]{%
		\centering
		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{FR_halfMH.png}}%
		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{latent_halfMH.png}}%
		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{A_halfMH.png}}%
	}
	\caption{half MH for latent, fit}
	\label{half MH: fit}
\end{figure}

Well, everything is stable now. But this implementation is too naive. Maybe use some more advanced adaptive sampling method?

%\begin{figure}[h!]
%	\centering
%	\includegraphics[width = .8\textwidth]{lds_blk_trace.png}
%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
%	\label{lds norms}
%\end{figure}\\
%Some results are in Figure \ref{fig:LDS labeled}. These are means from iteration 1000 to 2000.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{lds_blk_FR.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{lds_blk_latent.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{lds_blk_A.png}}%
%	}
%	\caption{LDS sample with labels}
%	\label{fig:LDS labeled}
%\end{figure}
%
%Personally, I think the fitting is not bad and the mixing is fine (in previous version, we may need 10000 iterations to get a not bad result. But here 2000 iterations did even better than before).

@@

@@

\textbf{I have deleted the old results to avoid confusion. I will post the new results here later.}

%This time I ran 1000 iterations. The trace of the samples is shown in Figure \ref{noA norms}.
%\begin{figure}[h!]
%	\centering
%	\includegraphics[width = .8\textwidth]{noA_blk_trace.png}
%	\caption{(Frobenius) norm for \(\mathbf{d}\), \(\mathbf{C}\), \(\mathbf{b}\) and \(\mathbf{A}\)}
%	\label{noA norms}
%\end{figure}\\
%
%Some results are in Figure \ref{fig:noA labeled}. These are means from iteration 500 to 1000.
%
%\begin{figure}[h!]
%	\makebox[\linewidth][c]{%
%		\centering
%		\subfigure[overall mean firing rate]{\label{fig:a}\includegraphics[width=0.55\textwidth]{noA_blk_FR.png}}%
%		\subfigure[latents]{\label{fig:b}\includegraphics[width=0.55\textwidth]{noA_blk_latent.png}}%
%		\subfigure[dynamics \(\mathbf{A}\)]{\label{fig:c}\includegraphics[width=0.55\textwidth]{noA_blk_A.png}}%
%	}
%	\caption{LDS sample with labels}
%	\label{fig:noA labeled}
%\end{figure}
%
%This time the mixing of loading (\(\mathbf{d}\) and \(\mathbf{C}\)) is not fine. That reminds me that \textbf{I may need to make loading within cluster related by MV-(Poisson)-GLM directly.}

\subsubsection{Unlabeled Data: Clustering}
\textbf{The clustering results are old. Update when the MFM is implemented}

Again, all results can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/results/gif}{folder}. The code can be found in this \href{https://github.com/weigcdsb/state-space-clustering/tree/main/LDS/blkDiag}{folder}.

@@

\section{TODO}
\begin{enumerate}
	\def\labelenumi{(\arabic{enumi})}
	\item
%	Think about MV-(Poisson)-GLM. Check \href{https://www.tandfonline.com/doi/full/10.1080/03610926.2012.743565?journalCode=lsta20}{ref1} and \href{https://www.tandfonline.com/doi/full/10.1080/02664763.2021.1877637?src=recsys}{ref2}.
	Do MH adaptively.
%	\item
%	After resolving (1), see if the MCMC is still trapped in local optima sometimes. Now, the chain seems will get stuck in local mode. Maybe updating the loading as a whole will remedy the problem a bit.
	\item
	Implement the mixture of finite mixture (MFM) by Jeff Miller.
	\item
	Find a more efficient way to generate new cluster parameters, otherwise the newly generated ones will always be rejected.
%	\item
%	improve DPMM. In current brute force implementation, the number of potential clusters can even go beyond number of neurons (\(N\)). There are several improvements, e.g. \href{https://link.springer.com/article/10.1007/s11222-009-9150-y}{Kalli et al., 2011}, \href{http://proceedings.mlr.press/v37/gea15.html}{Ge et al., 2015} and \href{https://link.springer.com/article/10.1007/s11222-014-9471-3}{Hastie et al, 2015}. Check them later.
	\item
	After all of them are resolved, switch to GP version
